Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-15 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Introduction.

Phospholipids containing various polar headgroups and acyl chains are essential building blocks of 
biological membranes. Lamellar phospholipid bilayer structures have been widely studied with various experimental 
and theoretical techniques as a simple model for cellular membranes~\cite{lipowsky95,tieleman97,klauda08,edholm08,tieleman10,piggot12,rabinovich13,marsh13}. 
Phospholipid molecules are composed of hydrophobic acyl chains which are connected by a glycerol backbone to a hydrophilic headgroup.
See Fig.~\ref{POPCstructure} for the structure of 1-palmitoyl-2-oleoylphosphatidylcholine (POPC).
The behaviour of the acyl chains in a lipid bilayer is relatively well understood~\cite{Israelachvili80,lipowsky95,tieleman97,klauda08,edholm08,tieleman10,marsh13}. 
The conformations sampled by the glycerol backbone and choline in a fluid bilayer are, however, not fully 
resolved since even the most accurate scattering and Nuclear Magnetic Resonance (NMR)
techniques give only a set of values that the structure has to fulfil, but
there is no unique way to derive the actual structure from them~\cite{seelig77b,skarjune79,Israelachvili80,jacobs80,davis83,strenk85,akutsu91,hong95b,hong96,semchyschyn04}.
Some structural details have been extracted from crystal structure, $^1$H NMR studies and Raman spectroscopy~\cite{hauser80,hauser81,hauser81b,akutsu81b,pascher92,hauser88,marsh06}
but general consensus about the structures sampled in the fluid state have not been reached~\cite{seelig77b,skarjune79,Israelachvili80,jacobs80,davis83,strenk85,hauser88,akutsu91,hong95b,hong96,semchyschyn04,marsh06}. 
On the other hand, the structural parameters for the glycerol backbone are similar for various biologically
relevant lipid species (phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG)) 
in various environments~\cite{gally81} and the structural parameters for the headgroup choline structures are similar in model membranes and
real cells (mouse fibroblast L-M cell)~\cite{scherer87}.
Thus, the solution of phosphatidylcholine glycerol backbone and choline structures would be 
useful for understanding a wide range of different biological membranes.

Classical atomistic molecular dynamics simulations have been widely used to study  
lipid bilayers~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12,rabinovich13}. As these models provide an atomistic
resolution description of the whole lipid molecule, they have potential to solve the glycerol backbone and 
headgroup structures. In particular, the experimental C--H bond order parameters for the glycerol backbone 
(g$_1$, g$_2$ and g$_3$) and choline ($\alpha$ and $\beta$) segments (see Fig.~\ref{POPCstructure} for definitions) are among the main parameters used in
attempts to derive lipid structures from experimental data~\cite{seelig77b,skarjune79,jacobs80,davis83,akutsu91,hong95b,semchyschyn04}.
These parameters are also routinely compared between experiments and simulations for the acyl chains~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12}.
Thus, the structures sampled in a simulation model that reproduces these and other experimental parameters, automatically
give an interpretation of the experiments. In other words they can be considered as reasonable atomistic resolution descriptions of
the behavior of lipid molecules in a bilayer.

The glycerol backbone and choline headgroup order parameters have been compared between simulations and experiments
in some studies~\cite{shinoda97,hogberg08,castro08,klauda10,kapla12,dickson12,poger12,ferreira13,chowdhary13,maciejewski14}, 
however much less frequently than for the acyl  chains~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12}.
The main reason is probably that the existing experimental data for the glycerol backbone
and choline headgroups is scattered over many publications and published in a format that is difficult to understand without some NMR expertise. 
In addition to the order parameters, also dihedral angles for glycerol backbone and headgroup estimated from experiments have been sometimes used to 
assess the quality of a simulation model~\cite{robinson94,essex94,kothekar96,hyvonen97,shinoda97,duong99}.

In this work we first review the most relevant experimental data for the glycerol backbone and choline headgroup order parameters
in a phosphatidylcholine lipid bilayer. Then the available atomistic resolution lipid models are carefully compared to the 
experimental data. The comparison reveals that the CHARMM36~\cite{klauda10}, GAFFlipid~\cite{dickson12} and MacRog models~\cite{maciejewski14}
have the most realistic glycerol backbone and choline structures. We also compare the glycerol backbone and choline 
structures between the most often used (Berger based) lipid model~\cite{berger97} 
%\todo{If there are no objections, we will leave the naming convention of different Berger versions as it is now.}
and the best performing models, to demonstrate that by using the 
order parameters we can distinguish the more reasonable structures from the less reasonable ones. However, none of the current models 
is accurate enough to properly resolve the atomistic resolution stuctures.

In addition to fully hydrated single component lipid bilayers, the glycerol backbone and choline order parameters
have been measured under a large number of different conditions. For example, as a function of hydration level~\cite{bechinger91,ulrich94,dvinskikh05b}, cholesterol content~\cite{brown78,ferreira13}
ion concentration~\cite{brown77,akutsu81,altenbach84,roux90,roux91}, temperature~\cite{gally75}, charged lipid content~\cite{roux90,roux91}, charged surfactant content~\cite{scherer89}, 
drug molecule concentration~\cite{browning82,kelusky84,castro08}, and protein content~\cite{roux89,kuchinka89} (listing only the publications most relevant for this work and the pioneering studies).
Awareness of the existence of this type of data allows the comparison of structural responses to varying conditions between simulations and experiments,
which can be used to validate the simulation models and to interpret the original experiments. 
In this publication we demonstrate the power of this approach for understanding the behaviour of a bilayer as a function of hydration level and cholesterol concentration.
Choline headgroup order parameters as function of ion concentration, and their relation to the ion binding affinity, are discussed elsewhere~\cite{ionpaper}.

>>>>

Phospholipids containing various polar headgroups and acyl chains are essential building blocks of 
biological membranes. Lamellar phospholipid bilayer structures have been widely studied with various experimental 
and theoretical techniques as a simple model for cellular membranes~\cite{lipowsky95,tieleman97,klauda08,edholm08,tieleman10,piggot12,rabinovich13,marsh13}. 
Phospholipid molecules are composed of hydrophobic acyl chains connected by a glycerol backbone to a hydrophilic headgroup;
see Fig.~\ref{POPCstructure} for the structure of 1-palmitoyl-2-oleoyl\-phosphatidyl\-choline (POPC).
The behaviour of the acyl chains in a lipid bilayer is relatively well understood~\cite{Israelachvili80,lipowsky95,tieleman97,klauda08,edholm08,tieleman10,marsh13}. 
The conformations sampled by the glycerol backbone and choline in a fluid bilayer are, however, not fully 
resolved as even the most accurate scattering and Nuclear Magnetic Resonance (NMR)
techniques give only a set of values that the structure has to fulfill, but
there is no unique way to derive the actual structure from them~\cite{seelig77b,skarjune79,Israelachvili80,jacobs80,davis83,strenk85,akutsu91,hong95b,hong96,semchyschyn04}.
Some structural details have been extracted from crystal structures, $^1$H NMR studies, and Raman spectroscopy~\cite{hauser80,hauser81,hauser81b,akutsu81b,pascher92,hauser88,marsh06},
but general consensus concerning the structures sampled in the fluid state has not been reached~\cite{seelig77b,skarjune79,Israelachvili80,jacobs80,davis83,strenk85,hauser88,akutsu91,hong95b,hong96,semchyschyn04,marsh06}. 
Importantly, the structural parameters for the glycerol backbone are similar for various biologically
relevant lipid species (phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG)) 
in various environments~\cite{gally81}, and the structural parameters for the choline headgroup are similar in model membranes and
real cells (mouse fibroblast L-M cell)~\cite{scherer87}.
Thus, resolving the PC-lipid glycerol and choline structures would be 
useful for understanding a wide range of different biological membranes.

Classical atomistic molecular dynamics simulations have been widely used to study  
lipid bilayers~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12,rabinovich13}. As these models provide an atomistic
resolution description of the whole lipid molecule, they have the potential to solve the glycerol backbone and 
headgroup structures. The experimental C--H bond order parameters
(routinely compared between experiments and simulations for the acyl chains~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12})
are also known for the glycerol backbone 
(g$_1$, g$_2$, and g$_3$) and choline ($\alpha$ and $\beta$) segments (see Fig.~\ref{POPCstructure} for definitions) and are among the main parameters used in
attempts to derive lipid structures from experimental data~\cite{seelig77b,skarjune79,jacobs80,davis83,akutsu91,hong95b,semchyschyn04}.
Notably, the structures sampled in a simulation that reproduces these %and other experimental
parameters will automatically comprise an interpretation of the experiments. In other words, such simulations can be considered as an accurate atomistic resolution description of
the behavior of lipid molecules in a bilayer.

Only a few studies~\cite{shinoda97,hogberg08,castro08,klauda10,kapla12,dickson12,poger12,ferreira13,chowdhary13,maciejewski14}
have compared
the glycerol backbone and choline headgroup order parameters between simulations and experiments. 
The main reason probably is that the existing experimental data for the glycerol backbone
and choline headgroups are scattered over many publications and published in a format that is difficult to understand without some NMR expertise. 
In addition to the order parameters, dihedral angles for the glycerol backbone and headgroup estimated from experiments have sometimes been used to 
assess the quality of a simulation model~\cite{robinson94,essex94,kothekar96,hyvonen97,shinoda97,duong99}.

In this work, we first review the most relevant experimental data for the glycerol backbone and choline headgroup order parameters
in a phosphatidylcholine lipid bilayer. Then the available atomistic resolution lipid models are carefully compared to the 
experimental data. The comparison reveals that the CHARMM36~\cite{klauda10}, GAFFlipid~\cite{dickson12}, and MacRog~\cite{maciejewski14} models
have the most realistic glycerol backbone and choline structures. We also compare the glycerol backbone and choline 
structures between the most often used (Berger-based) lipid model~\cite{berger97} 
%\todo{If there are no objections, we will leave the naming convention of different Berger versions as it is now.}
and the best performing models, to demonstrate that by using the 
order parameters we can distinguish the more reasonable structures from the less reasonable ones. However, none of the current models 
is accurate enough to properly resolve the atomistic resolution structures.

In addition to fully hydrated single component lipid bilayers, the glycerol backbone and choline order parameters
have been measured under a large number of changing conditions: hydration level~\cite{bechinger91,ulrich94,dvinskikh05b}, cholesterol content~\cite{brown78,ferreira13},
ion concentration~\cite{brown77,akutsu81,altenbach84,roux90,roux91}, temperature~\cite{gally75}, charged lipid content~\cite{roux90,roux91}, charged surfactant content~\cite{scherer89}, 
drug molecule concentration~\cite{browning82,kelusky84,castro08}, and protein content~\cite{roux89,kuchinka89} (listing only the publications most relevant for this work and the pioneering studies).
Existence of these data allows the comparison of structural responses to varying conditions between simulations and experiments,
in other words, validation of the simulation models and interpretation of the original experiments. 
Here we demonstrate the power of this approach in understanding the behaviour of a bilayer as a function of hydration level and cholesterol content.
Choline headgroup order parameters as function of ion concentration, and their relation to the ion binding affinity, are discussed elsewhere~\cite{ionpaper}.
